---
output:
  pdf_document: default
  html_document: default
---
Required Packages and settings: 
```{r echo = TRUE}
library(tidyverse)
library(data.table)
library(ggplot2)
library(dplyr)
library(xtable)
```

Reading in Data, I'll assume that you have set your Working Directory, and the 
file downloaded accordingly off the web. 

```{r echo=TRUE}
StormData <- read.csv("Data/repdata_data_StormData.csv.bz2"
                      , header = TRUE
                      , sep = ",")
```

What it do? 

```{r echo=TRUE}
summary(StormData)
str(StormData)
head(StormData)
```

So 38 different columns (variables), and my suspicion is that we will not end up
utilizing all of them. For the sake of saving the CPU some work, lets go ahead 
and condense this dataframe. 

First we are ensuring that we are getting a non-zero number in atleast one 
of the "fatalities, injuries, property damage, and crop damage" categories.
After all, if they are not causing measurable damage they should not show
up in our findings.

Second, there is some cleaning we should do on the date types to make sure that 
we are getting accurate data on the years that events are taking place. The 
documentation alludes the distribution of weather events are heavily skewed 
towards the later years, since the records are spare. Even though there is a 
clear right skew to the data in regards to year I believe this problem is
not worth a subset. 

Third, we need to look at the veracity of the "EVTYPES" column, to make sure
the events are properly recorded. According to the documentation, there are 
only 47 event types. Lets see what we have: 

```{r echo = TRUE}
CleanStormData <- subset(StormData, (FATALITIES > 0 | INJURIES > 0 | 
                                    PROPDMG > 0 | CROPDMG > 0)
                                  , select =c("BGN_DATE"
                                             ,"STATE"
                                             ,"EVTYPE"
                                             ,"FATALITIES"
                                             ,"INJURIES"
                                             ,"PROPDMG"
                                             ,"PROPDMGEXP"
                                             ,"CROPDMG"
                                             ,"CROPDMGEXP"))

CleanStormData$year <- as.numeric(format(
                                  as.Date(CleanStormData$BGN_DATE
                                        , format = "%m/%d/%Y %H:%M:%S"),"%Y"))

hist(CleanStormData$year, breaks = 10)
CleanStormData$EVTYPE <- toupper(CleanStormData$EVTYPE)
```


Now, we're running into our second problem, the Property Damage and Crop Damage
columns. These documentation suggests that we get the following: 

Symbol      Magnitude 
"B"         Billion 
"M"         Million 
"K"         Thousand 

Simple enough. Lets see what we get. 
```{r echo=TRUE}
table(CleanStormData$PROPDMGEXP) 
table(CleanStormData$CROPDMGEXP)
```

Alright, so there are a few other characters listed, including blanks 

I'm making a few assumptions here: 

    1.) The capitalization is user generated and has no "real" difference. A "M" 
        will be treated the same as "m"

    2.) The "H"/"h" character symbolizes a magnification of 100 

    3.) The "-", "+", "0" "?" characters are missing value characters and will 
        have magnification of 1 (identity property)

    4.) The numeric "2", "3", etc. yield the appropriate 10^X magnification, 
        where X = "2", X = "3" respectively. 

With those assumptions in mind, here is the cleaning for the Economic devastation

```{r echo=TRUE}
head(toupper(CleanStormData$PROPDMGEXP),3)
head(toupper(CleanStormData$CROPDMGEXP),3)

Magnification <- function(exp) {
    exp <- toupper(exp);
    if (exp == "")  return (10^0);
    if (exp == "-") return (10^0);
    if (exp == "?") return (10^0);
    if (exp == "+") return (10^0);
    if (exp == "0") return (10^0);
    if (exp == "1") return (10^1);
    if (exp == "2") return (10^2);
    if (exp == "3") return (10^3);
    if (exp == "4") return (10^4);
    if (exp == "5") return (10^5);
    if (exp == "6") return (10^6);
    if (exp == "7") return (10^7);
    if (exp == "8") return (10^8);
    if (exp == "9") return (10^9);
    if (exp == "H") return (10^2);
    if (exp == "K") return (10^3);
    if (exp == "M") return (10^6);
    if (exp == "B") return (10^9);
    return (NA);
}

CleanStormData$TotalPropCost <- with(CleanStormData, as.numeric(PROPDMG) 
                               * sapply(PROPDMGEXP, Magnification))/10^9

CleanStormData$TotalCropCost <- with(CleanStormData, as.numeric(CROPDMG) 
                              * sapply(CROPDMGEXP, Magnification))/10^9

```

Our final step before actually plotting and visualizing our results is making 
sure that our data is properly aggregated (by sum) and ordered. 

```{r echo=TRUE}

#Impact on Public Health
Fatalities <- aggregate(x = list(Impact = CleanStormData$FATALITIES)
                               , by = list(EVENT_TYPE = CleanStormData$EVTYPE)
                               , FUN = sum)

Fatalities <- Fatalities[order(Fatalities$Impact, 
                               decreasing = TRUE),]

# Impact on the Economy 
PropertyCost <- aggregate(x = list(Impact = CleanStormData$TotalPropCost)
                                 , by = list(EVENT_TYPE= CleanStormData$EVTYPE)
                                 , FUN = sum) 

PropertyCost <- PropertyCost[order(PropertyCost$Impact,
                                   decreasing = TRUE),]

CropCost <- aggregate(x = list(Impact = CleanStormData$TotalCropCost)
                              , by = list(EVENT_TYPE = CleanStormData$EVTYPE)
                              , FUN = sum) 

CropCost <- CropCost[order(CropCost$Impact, 
                           decreasing = TRUE),]
```

At this point, we are finally ready to visualize our results! 

```{r echo=TRUE}

print(xtable(head(PropertyCost, 10),
             caption = "Top 10 Weather Events Most Harmful to Property"),
             caption.placement = 'top',
             type = "html",
             include.rownames = FALSE,
             html.table.attributes='class="table-bordered", width="100%"')


print(xtable(head(CropCost, 10),
             caption = "Top 10 Weather Events Most Harmful to Crops"),
             caption.placement = 'top',
             type = "html",
             include.rownames = FALSE,
             html.table.attributes='class="table-bordered", width="100%"')

print(xtable(head(Fatalities, 10),
             caption = "Top 10 Deadliest Weather Events"),
             caption.placement = 'top',
             type = "html",
             include.rownames = FALSE,
             html.table.attributes='class="table-bordered", width="100%"')

```

First we are going to visualize the most damaging events to the US economy, the 
property and weather related costs. After picking the "top 10" most dangerous 
events in each category. Floods and droughts are bad. 

```{r}

CropCostPlot <- ggplot(head(CropCost,10)
                      , aes(x = EVENT_TYPE
                          , y = Impact)) +
                geom_bar(stat = "Identity") + 
                coord_flip() +
                xlab("Event Type") + 
                ylab("Total Crop Damage in USD (Billions)") + 
                theme(plot.title = element_text(size = 14
                                                , hjust = 0.5)) + 
                ggtitle("Top 10 Weather Events Most Impactful to Crops")

PropertyCostPlot <- ggplot(head(PropertyCost,10)
                      , aes(x = EVENT_TYPE
                          , y = Impact)) +
                geom_bar(stat = "Identity") + 
                coord_flip() +
                xlab("Event Type") + 
                ylab("Total Crop Damage in USD (Billions)") + 
                theme(plot.title = element_text(size = 14
                                                , hjust = 0.5)) + 
                ggtitle("Top 10 Weather Events Most Impactful to Property")

print(CropCostPlot)
print(PropertyCostPlot)
```

```{r}

FatalitiesPlot <- ggplot(head(Fatalities,10)
                      , aes(x = EVENT_TYPE
                          , y = Impact)) +
                geom_bar(stat = "Identity") + 
                coord_flip() +
                xlab("Event Type") + 
                ylab("Aggregated Deaths") + 
                theme(plot.title = element_text(size = 14
                                                , hjust = 0.5)) + 
                ggtitle("Top 10 Deadliest Weather Events")


print(FatalitiesPlot)
```
Who Knew? Tornadoes = Bad 