Required Packages and settings: 
```{r}
echo = TRUE
library(sqldf)
library(data.table)
library(ggplot2)
library(dplyr)
```
Reading in Data
```{r}
StormData <- read.csv("Data/repdata_data_StormData.csv.bz2"
                      , header = TRUE
                      , sep = ",")
```

What it do? 
```{r echo=TRUE}
summary(StormData)
str(StormData)
head(StormData)
```

Okay, cool... There is a good bit of cleaning that we should do on this to
do the data justice. For our purposes of answering two questions 1.)
What sort of events are most dangerous to population health and 2.) What
sort of events are most economically expensive; we do not require a deep level 
of cleaning. 

Lets see what we got by year: 
```{r}
if (dim(StormData)[2] == 37) {
    StormData$year <- as.numeric(format(
                                as.Date(
                                  StormData$BGN_DATE
                                , format = "%m/%d/%Y %H:%M:%S"), "%Y"))
}
hist(StormData$year, breaks = 30)
```

```{r}
GoodStormData <- StormData[StormData$year >= 1990, ]
```

Now it looks like we're in the position of subsetting by year, and evaluating
the different impacts of storms in regards to population health: fatalities and 
injuries. 
```{r}

TopTen <- function(fieldName, top = 10, dataset = GoodStormData) {
    index <- which(colnames(dataset) == fieldName)
    field <- aggregate(dataset[, index], by = list(dataset$EVTYPE), FUN = "sum")
    names(field) <- c("EVTYPE", fieldName)
    field <- arrange(field, field[, 2], decreasing = T)
    field <- head(field, n = top)
    field <- within(field, EVTYPE <- factor(x = EVTYPE, levels = field$EVTYPE))
    return(field)
}

#health 
fatalities <- TopTen("FATALITIES", dataset = GoodStormData)
injuries <- TopTen("INJURIES", dataset = GoodStormData)

```

